Disk Heating, Galactoseismology, and the Formation of Stellar Halos

### 2017-06-26 paper/paper.tex
minor edits, whitespace

\begin{figure}[t]
\label{fig:ting}
\centering
%\includegraphics[width=3 cm]{logo-mdpi}
\caption{\label{fig:ting}
Summary of spatial distributions of M giants
The distance measurements are from photometry alone and have expected absolute uncertainties of $\approx$20\% \cite{sheffield14}.
}
\end{figure}
% APW COMMENT: Maybe Ting would be willing to share her data, and I can re-make
% the figure? You know how I am about visualizations... :)
% KVJ: that makes me very happy

@@

In subsequent work we continued the spectroscopic survey to look at A13, the new group M giant discovered through the group-finding algorithm \cite{sharma10}. A13 overlaps the TriAnd clouds in Galactic longitude, but is apparent in the Northern Galactic hemisphere at slightly brighter magnitudes. The spectra showed that, like the TriAnd clouds, this  group had a coherent velocity structure with low dispersion and a steady gradient with $l$, confirming the genuine association of its members \cite{li17}.
% APW COMMENT: (I'm sure you know) but this paragraph needs more detail about
% the Li et al. paper. Adding this as a TODO for us both.
% KVJ: I just added a couple of sentences. Don't want seciton too long
...

\begin{figure}[t]
\centering
%\includegraphics[width=3 cm]{logo-mdpi}
\caption{\label{fig:ting_vel}
Summary of velocity distributions of M giants}
\end{figure}

@@

\begin{figure}[t]
\centering
%\includegraphics[width=3 cm]{logo-mdpi}
\caption{\label{fig:apw}
v vs l for M giants vs RR Lyrae}
\end{figure}

@@

The origin of stellar associations can also be explored through the chemical compositions of their constituent stars.
However, our finding (described in Section \ref{sec:populations}) that the low-latitude-structures all have stellar populations (as indicated by $f_{\rm RR:M}$) that look more like the disk than known Galactic satellites is at odds with prior work on abundance patterns.
For example, high-resolution spectra of 21 M giants in Mon/GASS \cite{chou2010b} showed 
[Ti/Fe] lower by up to 0.4 dex compared to the mean trends known for main-sequence stars of the Galactic disk (e.g Reddy et al. 2003, Bensby et al. 2014) and a mean offset for [Y/Fe] of about 0.2 dex at [Fe/H] $\approx -0.5$. A comparison with results from the same group for stars in the Sgr dwarf spheroidal galaxy \cite{chou2010a} suggested more similar patterns to those seen in Mon/GASS and support  an external origin for this structure. Subsequent work comes to similar conclusions for the TriAnd Clouds \cite{chou11}.
%It should be noted, however, that their results do not agree with the abundances of Ti and, in particular, La, measured in the Canis Major overdensity by \cite{sbordone2005}. Their analysis was based on spectra covering a limited spectral region in the near-IR, including 11 neutral iron and 2 neutral titanium lines. Also \cite{chou2010b} caution that the effects of non-local thermodynamic equilibrium (non-LTE) may be significant, since their measurement of metallicity and Ti abundances relies on the LTE analysis of lines of neutral species,

In order to explore the apparent contradiction between the stellar populations and abundance work we have recently obtained high-resolution and signal-to-noise spectra of 15 stars in TriAnd 1 and A13 overdensities. 14 stars were observed with the HIRES-S spectrograph at the Keck-1 telescope \cite{vogt1994} and 1 star was observed using the the UVES spectrograph at the VLT (Program ID: 097.B-0770A). 
%The spectral resolving power R of the HIRES spectra is 36\,000 and the UVES data have R $\sim$ 47\,000. All Keck spectra cover the full optical region, from $4800$ to $8770$ $\AA$, although the exact wavelength coverage varies as several slightly different instrument configurations were tried in an effort to get all the key lines into a single exposure. The signal-to-noise (SNR)/\AA of the HIRES spectra near 5200 \AA in the continuum at the center of the echelle order exceeds 200. For the UVES spectrum, the SNR of 50 near 5500 \AA was achieved. The fundamental parameters and chemical abundances of stars were determined using 2MASS and APASS photometry and the high-resolution spectra. All 15 stars are M-type giants with effective temperatures Teff ~ 3800 K and surface gravities log(g) ~ 1 dex. The $T_{\rm eff}$ estimates were derived using the Infrared Flux Method (Casagrande et al. 2010, 2014). Chemical abundances were measured for 6 chemical elements, including O, Na, Mg, Ca, Fe, Eu using LTE and the standard MARCS stellar model atmospheres (Gustafsson et al. 2008). We have also performed detailed analysis of the effects of non-local thermodynamic equilibrium (non-LTE) for the elements, where detailed calculations are available \cite{bergemann2011, bergemann2012, bergemann2016}, however, the NLTE corrections for the chosen lines are minor and do not change our conclusions. In what follows, we use our LTE results, because all studies of chemical abundances in dSph systems and most studies in the Milky Way to date have employed LTE with 1D hydrostatic models
We find that stars in the TriAnd 1 and in A13 overdensities have extremely similar chemical abundances, with the abundance dispersion $\leq 0.06$ dex for most chemical elements. The TriAnd 1 and A13 stars also have a very narrow metallicity spread,  [Fe/H] $= -0.92 \pm 0.06$ dex. The abundances of all measured alpha-elements are uniformly enhanced at the level of 0.43 $\sim$ 0.06 dex for [O/Fe], 0.41 $\sim$ 0.02 dex for [Mg/Fe], $0.35 \pm 0.06$ dex for [Na/Fe], and 0.15 $\sim$ 0.03 dex for [Eu/Fe] (Figure 1). At this level, the TriAnd 1 and A13 stars are consistent with the abundances of the in-situ formed Milky Way thick disk stars \cite{fuhrmann2002,bergemann2014,bensby2014}.
\footnote{The only exception is [Ca/Fe], which is solar with the dispersion of 0.06 dex; however, the NLTE abundance correction for the \ion{Ca}{i} lines for M giants is 0.1 dex that will bring the results in agreement with the MW disk studies that are based on FGK dwarfs, for which the NLTE correction is much smaller \cite{Merle2011}}. 
The abundance ratios are too high for the chemical abundance patterns observed in the stars of the Galactic satellites (Bonifacio et al. 2010, Shetrone et al. 2001, 2003, Tolstoy et al. 2009, de Boer et al. 2014), which are known to have low, typically solar ( [O/Fe], [Mg/Fe]) or even sub-solar ([Na/Fe]), ratios at [Fe/H] $\sim -1$. 
%For example, Fornax, a dSph galaxy most close to TriAnd 1 and A13 in metallicity, has [Na/Fe] $\sim -0.6$ dex and [Mg/Fe] $\sim 0$ at [Fe/H] $=-1$ (Letarte et al. 2010, Kirby 2010, Lemasle et al. 2014) \footnote{Note that to avoid systematic differences between abundance measurements obtained from spectra taken with different instruments, we have chosen here to compare with the data taken with the same instrument; here UVES and FLAMES at the VLT, ESO, or with HIRES at Keck.}.

@@

Overall, comparing our work to abundances of disk stars and satellite populations obtained using analogous data sets and reduction techniques, our results suggest that the birthplace of stars in TriAnd 1 and A13 was the outer Galactic disk rather than an infalling satellite galaxy.
A paper describing this work is currently in preparation (Bergemann et al, 2017).
%it is unlikely that the TriAnd 1 and A13 stars originate from a disrupted dSph galaxy

@@

Now that we have a physical picture of the origin of such features, as well as prospects for mapping them further within the Milky Way and detecting analogous substructures around other galaxies, we can move on to discussing how useful they are as probes of dynamics and history.
It is important to have these uses in mind to motivate and frame  future work.
While the mere existence of these structures is interesting, they contain a tiny fraction of the stars in galaxies spread out over a large area --- these properties naturally make them difficult to map, either because their unique signatures can be lost in the foreground star counts (e.g. in the Milky Way) , or  because the required surface brightness limits for detection are prohibitively low (for integrated light).

@@

Below are just three examples of where these structures could promise new insights into some classic questions.
\begin{description}
\item{\it Disk heating mechanisms ---}
It has been understood for a long time that disks can evolve significantly due to mergers, major or minor, and hence that their current structures bear witness to their accretion history \cite{toth92,quinn93,walker96,velazquez99}.
This understanding has fueled a significant literature on the importance of the heating of galactic disks in response to encounters with other dark matter halos (that may or may not contain their own galaxies) \cite{font01,ardi03,benson04,stewart08,hopkins08,villalobos08,purcell09,kazantzidis09,sachdeva16,moetazedian16}.
In general, these studies have concentrated on the overall effects of many encounters on global properties, such as the thickness and vertical velocity dispersion in disks.
Their results have traditionally been compared to these scales in samples of galaxies.
In contrast, the identification and mapping of vertical waves in our Milky Way associated with ongoing interactions gives us the opportunity to dissect an individual disk heating event in progress.
We can use this to check our understanding of the mechanism directly and in detail rather than assessing its importance through collective effects and longterm, phase-mixed signatures.
\item{\it Stellar halo formation processes ----}
The last decade has seen increasing interest in assessing how much of the content of stellar halos could be made from stars originally formed in the disks of the galaxies that they surround. Hydrodynamical simulations of galaxy formaion suggest that tens of percent of the inner halo might be formed this way
\cite{abadi06,zolotov09,zolotov10,font11,mccarthy12,tissera13,tissera14,pillepich15,cooper15}.
Preliminary arguments for the existence of this ``kicked-out-disk'' population were based on transitions in the density or orbital structures of stellar halos \cite[e.e.][]{carollo07}.
However, such transitions were also found to naturally occur in purely-accreted models of stellar halos \cite{deason13}.
More convincing observational evidence for disk stars in the halo is just beginning to emerge through studies with look for stars with halo-like kinematics, but disk-like abundances around M31 \cite{dorman13} and the Milky Way \cite{sheffield12,hawkins15,bonaca17}.
Our own work adds new perspectives on this stellar halo formation process with the detection and modeling of disk stars that may be in transition from the disk to the halo.
\item{\it Galactoseismic probes of interactions and dark matter ---}
The response of a disk to an encounter will depend on its own properties, the properties of the dark matter halo in which it is embedded and the mass and orbit of the perturbing satellite.
This leads to the suggestion that, analogous to helioseismic investigations of the structure of our Sun, maps of a disk response --- such as those described in Section \ref{sec:obs} for our Milky Way --- might be similarly inverted to tell us about (e.g.) the structure of the dark matter halo
\cite{widrow12}.
Indeed, recent studies of the signatures of encounters in the very outskirts of extended HI disks have successfully used simulations combined with an analytic understanding to find how the observed characteristics of the disturbed gas can be simply related to properties of the perturbing object \cite{chakrabarti09,chakrabarti11b,chang11}.
\end{description}

Uplifitng/Concluding sentence.

>>>>

\begin{figure}[!ht]
\label{fig:ting}
\centering
\includegraphics[width=6 in]{figures/xy_Rz.pdf}
\caption{\label{fig:ting}
Summary of the spatial distribution of M giants in each of the three
low-latitude structures.
Note that at lower Galactic latitudes, the lack of candidate M giant members is
due to selection effects and crowding near the midplane; we expect the
structures to continue towards even lower latitudes, but blend with ordinary
disk stars.
Markers represent individual stars identified as likely members of each of the
three structures discussed in this work (see figure legend).
Distance estimates come from photometric alone and have expected absolute
uncertainties around $\approx$20\% for TriAnd and A13 \cite{sheffield14,li17}
and $\approx$25\% for GASS.
Grey curves in left panel show Galactocentric circles with cylindrical radii,
$R$, indicated on the figure.
The position of the sun is marked with the solar symbol $\odot$.
}
\end{figure}

@@

In subsequent work we continued the spectroscopic survey to look at A13, the new group M giant discovered through the group-finding algorithm \cite{sharma10}. A13 overlaps the TriAnd clouds in Galactic longitude, but is apparent in the Northern Galactic hemisphere at slightly brighter magnitudes. The spectra showed that, like the TriAnd clouds, this  group had a coherent velocity structure with low dispersion and a steady gradient with $l$, confirming the genuine association of its members \cite{li17}.

\begin{figure}[!ht]
\centering
\includegraphics[width=5 in]{figures/vgsr.pdf}
\caption{\label{fig:ting_vel}
Summary of the velocity distribution of M giants in each of the three
low-latitude structures.
Markers represent individual stars identified as likely members of each of the
three structures discussed in this work (see figure legend).
Grey curves show circular orbits in the Galactic disk midplane with circular
velocity equal to $220~\kms$ at several Galactocentric cylindrical radii, $R$,
as indicated on the figure.
Velocity uncertainties are typically the same as or smaller than the marker
sizes.
}
\end{figure}

@@

\begin{figure}[t]
\centering
\includegraphics[width=6 in]{figures/triand_rrlyrae}
\caption{\label{fig:apw}
Comparison of the velocity distribution for M giants in the TriAnd structure
(top panel) with velocities for RR Lyrae stars in the same region of sky and
distance range (bottom panel).
}
\end{figure}

@@

The origin of stellar associations can also be explored through the chemical compositions of their constituent stars.
However, our finding (described in Section \ref{sec:populations}) that the low-latitude-structures all have stellar populations (as indicated by $f_{\rm RR:M}$) that look more like the disk than known Galactic satellites is at odds with prior work on abundance patterns.
For example, high-resolution spectra of 21 M giants in Mon/GASS \cite{chou2010b} showed
[Ti/Fe] lower by up to 0.4 dex compared to the mean trends known for main-sequence stars of the Galactic disk (e.g., Reddy et al. 2003, Bensby et al. 2014) and a mean offset for [Y/Fe] of about 0.2 dex at ${\rm [Fe/H]} \approx -0.5$. A comparison with results from the same group for stars in the Sgr dwarf spheroidal galaxy \cite{chou2010a} suggested more similar patterns to those seen in Mon/GASS and support  an external origin for this structure. Subsequent work comes to similar conclusions for the TriAnd Clouds \cite{chou11}.
% APW COMMENT: there are hand-typed references in the paragraph above! We need to add to the bibliography

%It should be noted, however, that their results do not agree with the abundances of Ti and, in particular, La, measured in the Canis Major overdensity by \cite{sbordone2005}. Their analysis was based on spectra covering a limited spectral region in the near-IR, including 11 neutral iron and 2 neutral titanium lines. Also \cite{chou2010b} caution that the effects of non-local thermodynamic equilibrium (non-LTE) may be significant, since their measurement of metallicity and Ti abundances relies on the LTE analysis of lines of neutral species,

In order to explore the apparent contradiction between the stellar populations and abundance work we have recently obtained high-resolution and signal-to-noise spectra of 15 stars in TriAnd 1 and A13 overdensities. 14 stars were observed with the HIRES-S spectrograph at the Keck-1 telescope \cite{vogt1994} and 1 star was observed using the the UVES spectrograph at the VLT (Program ID: 097.B-0770A).
%The spectral resolving power R of the HIRES spectra is 36\,000 and the UVES data have R $\sim$ 47\,000. All Keck spectra cover the full optical region, from $4800$ to $8770$ $\AA$, although the exact wavelength coverage varies as several slightly different instrument configurations were tried in an effort to get all the key lines into a single exposure. The signal-to-noise (SNR)/\AA of the HIRES spectra near 5200 \AA in the continuum at the center of the echelle order exceeds 200. For the UVES spectrum, the SNR of 50 near 5500 \AA was achieved. The fundamental parameters and chemical abundances of stars were determined using 2MASS and APASS photometry and the high-resolution spectra. All 15 stars are M-type giants with effective temperatures Teff ~ 3800 K and surface gravities log(g) ~ 1 dex. The $T_{\rm eff}$ estimates were derived using the Infrared Flux Method (Casagrande et al. 2010, 2014). Chemical abundances were measured for 6 chemical elements, including O, Na, Mg, Ca, Fe, Eu using LTE and the standard MARCS stellar model atmospheres (Gustafsson et al. 2008). We have also performed detailed analysis of the effects of non-local thermodynamic equilibrium (non-LTE) for the elements, where detailed calculations are available \cite{bergemann2011, bergemann2012, bergemann2016}, however, the NLTE corrections for the chosen lines are minor and do not change our conclusions. In what follows, we use our LTE results, because all studies of chemical abundances in dSph systems and most studies in the Milky Way to date have employed LTE with 1D hydrostatic models
We find that stars in the TriAnd 1 and in A13 overdensities have extremely similar chemical abundances, with the abundance dispersion $\leq 0.06$ dex for most chemical elements. The TriAnd 1 and A13 stars also have a very narrow metallicity spread,  ${\rm [Fe/H]} = -0.92 \pm 0.06$ dex. The abundances of all measured alpha-elements are uniformly enhanced at the level of 0.43 $\sim$ 0.06 dex for [O/Fe], 0.41 $\sim$ 0.02 dex for [Mg/Fe], $0.35 \pm 0.06$ dex for [Na/Fe], and 0.15 $\sim$ 0.03 dex for [Eu/Fe]. At this level, the TriAnd 1 and A13 stars are consistent with the abundances of the in-situ formed Milky Way thick disk stars \cite{fuhrmann2002,bergemann2014,bensby2014}.
\footnote{The only exception is [Ca/Fe], which is consistent with solar with the dispersion of 0.06 dex; however, the NLTE abundance correction for the \ion{Ca}{i} lines for M giants is 0.1 dex that will bring the results in agreement with the MW disk studies that are based on FGK dwarfs, for which the NLTE correction is much smaller \cite{Merle2011}}.
The abundance ratios are too high for the chemical abundance patterns observed in the stars of the Galactic satellites (Bonifacio et al. 2010, Shetrone et al. 2001, 2003, Tolstoy et al. 2009, de Boer et al. 2014), which are known to have low, typically solar ( [O/Fe], [Mg/Fe]) or even sub-solar ([Na/Fe]), ratios at [Fe/H] $\sim -1$.
% APW COMMENT: there are hand-typed references in the paragraph above! We need to add to the bibliography

%For example, Fornax, a dSph galaxy most close to TriAnd 1 and A13 in metallicity, has [Na/Fe] $\sim -0.6$ dex and [Mg/Fe] $\sim 0$ at [Fe/H] $=-1$ (Letarte et al. 2010, Kirby 2010, Lemasle et al. 2014) \footnote{Note that to avoid systematic differences between abundance measurements obtained from spectra taken with different instruments, we have chosen here to compare with the data taken with the same instrument; here UVES and FLAMES at the VLT, ESO, or with HIRES at Keck.}.

@@

Overall, comparing our work to abundances of disk stars and satellite populations obtained using analogous data sets and reduction techniques, our results suggest that the birthplace of stars in TriAnd 1 and A13 was the outer Galactic disk rather than an infalling satellite galaxy.
A paper describing this work is currently in preparation (Bergemann et al., in prep.).

% APW COMMENT: there are hand-typed references in the paragraph above! We need to add to the bibliography

%it is unlikely that the TriAnd 1 and A13 stars originate from a disrupted dSph galaxy

@@

Now that we have a physical picture of the origin of such features, as well as prospects for mapping them further within the Milky Way and detecting analogous substructures around other galaxies, we can move on to discussing how useful they are as probes of dynamics and history.
It is important to have these uses in mind to motivate and frame  future work.
While the mere existence of these structures is interesting, they contain a tiny fraction of the stars in galaxies spread out over a large area --- these properties naturally make them difficult to map, either because their unique signatures can be lost in the foreground star counts (e.g., in the Milky Way), or  because the required surface brightness limits for detection are prohibitively low (for integrated light).

@@

Below are just three examples of where these structures could promise new insights into some classic questions.
\begin{description}
\item{\it Disk heating mechanisms ---}
It has been understood for a long time that disks can evolve significantly due to mergers, major or minor, and hence that their current structures bear witness to their accretion history \cite{toth92,quinn93,walker96,velazquez99}.
This understanding has fueled a significant literature on the importance of the heating of galactic disks in response to encounters with other dark matter halos (that may or may not contain their own galaxies) \cite{font01,ardi03,benson04,stewart08,hopkins08,villalobos08,purcell09,kazantzidis09,sachdeva16,moetazedian16}.
In general, these studies have concentrated on the overall effects of many encounters on global properties, such as the thickness and vertical velocity dispersion in disks.
Their results have traditionally been compared to these scales in samples of galaxies.
In contrast, the identification and mapping of vertical waves in our Milky Way associated with ongoing interactions gives us the opportunity to dissect an individual disk heating event in progress.
We can use this to check our understanding of the mechanism directly and in detail rather than assessing its importance through collective effects and longterm, phase-mixed signatures.
\item{\it Stellar halo formation processes ----}
The last decade has seen increasing interest in assessing how much of the content of stellar halos could be made from stars originally formed in the disks of the galaxies that they surround. Hydrodynamical simulations of galaxy formaion suggest that tens of percent of the inner halo might be formed this way
\cite{abadi06,zolotov09,zolotov10,font11,mccarthy12,tissera13,tissera14,pillepich15,cooper15}.
Preliminary arguments for the existence of this ``kicked-out-disk'' population were based on transitions in the density or orbital structures of stellar halos \cite[e.g.,][]{carollo07}.
However, such transitions were also found to naturally occur in purely-accreted models of stellar halos \cite{deason13}.
More convincing observational evidence for disk stars in the halo is just beginning to emerge through studies with look for stars with halo-like kinematics, but disk-like abundances around M31 \cite{dorman13} and the Milky Way \cite{sheffield12,hawkins15,bonaca17}.
Our own work adds new perspectives on this stellar halo formation process with the detection and modeling of disk stars that may be in transition from the disk to the halo.
\item{\it Galactoseismic probes of interactions and dark matter ---}
The response of a disk to an encounter will depend on its own properties, the properties of the dark matter halo in which it is embedded and the mass and orbit of the perturbing satellite.
This leads to the suggestion that, analogous to helioseismic investigations of the structure of our Sun, maps of a disk response --- such as those described in Section \ref{sec:obs} for our Milky Way --- might be similarly inverted to tell us about (e.g.) the structure of the dark matter halo
\cite{widrow12}.
Indeed, recent studies of the signatures of encounters in the very outskirts of extended HI disks have successfully used simulations combined with an analytic understanding to find how the observed characteristics of the disturbed gas can be simply related to properties of the perturbing object \cite{chakrabarti09,chakrabarti11b,chang11}.
\end{description}

Uplifting/Concluding sentence.
